---
title: "Legends"
author: "Valencia"
toc: true
number-sections: true
format: 
  html: 
   embed-resources: true
---

## Basic Box Plot
 - `library(ggplot2)`: To load `ggplot` the function to process on R Script
 
 - `ggplot(PlantGrowth, aes(x = group, y = weight))` Use `PlantGrowth` dataset and set `group` on the x-axis, and `weight` on the y-axis.

```{r}
library(ggplot2)
ggplot(PlantGrowth, aes(x = group, y = weight)) +
  geom_boxplot() +
  labs(title = 'Basic Box Plot',
       caption = 'Valencia, DV, THU 2024') +
  theme(plot.title = element_text(hjust = 0.5))
```

## Add Legends
 - `library(gridExtra)`: To load the `gridextra` to use the `grid.arrange` function
 - `fill`: To add colors inside the charts
 - `color`: To add color of the outline of the charts
 - `grid.arrange`: To arrange the boxplots
 - `top`: To add title on the top of the plot
 - `bottom`: To add captions at the bottom of the plot

```{r}
library(gridExtra)
p1 <- ggplot(PlantGrowth, aes(x = group, y = weight)) + 
  geom_boxplot()
p2 <- ggplot(PlantGrowth, aes(x = group, 
                              y = weight, 
                              fill = group)) + 
  geom_boxplot()
p3 <- ggplot(PlantGrowth, aes(x = group, 
                              y = weight,
                              color = group)) + 
  geom_boxplot()

grid.arrange(p1,p2,p3,ncol=3,
             top='Add Legend', 
             bottom = 'Valencia, DV, THU 2024')

```

## Remove Legend
 - `guides(fill = FALSE)` is to remove the Legend

```{r}
ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) + 
  geom_boxplot() +
  guides(fill = FALSE) +
  ggtitle('Remove Legend') +
  labs(caption = 'Valencia, DV, THU 2024') +
  theme(plot.title = element_text(hjust=0.5))
```

## Change Color of Legend
 - `scale_fill_brewer(palette = "Pastel2")` is to change the color/theme of the Legend

```{r}
ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  ggtitle('Change Color of Legend') +
  labs(caption = 'Valencia, DV, THU 2024') +
  theme(plot.title = element_text(hjust=0.5))
```

## Change Position of Legend

 - `legend.position = 'top'` is to change the position of the legend to the top
 
```{r}
ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  ggtitle('Change Color of Legend') +
  labs(caption = 'Valencia, DV, THU 2024') +
  theme(plot.title = element_text(hjust=0.5),
        legend.position = 'top')
```

## Practice
 - `p2 <- p1 + theme(legend.position = 'bottom')` is to change legend position to top, bottom, left, and right

```{r}
p1 <- ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Pastel2") +
  ggtitle('Change Color of Legend') +
  labs(caption = 'Valencia, DV, THU 2024') +
  theme(plot.title = element_text(hjust=0.5),
        legend.position = 'top')
p2 <- p1 + theme(legend.position = 'bottom')
p3 <- p1 + theme(legend.position = 'left')
p4 <- p1 + theme(legend.position = 'right')

grid.arrange(p1, p2, p3, p4, ncol=2,
             top='Chainging Position of Legend')
```

